---
title: "TIMES SERIES"
author: "Cheikh Mbacké BEYE"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      comment = NA,
                      warning = NA,
                      fig.width=12,
                      fig.height = 6
                      )
```
<style>
div.blue pre {background-color:#ccffcc;}
div.blue pre.r{background-color:#ccffff;}

</style>
<div class='blue'>

```{r}
library(ggplot2)
library(astsa)
library(xts)
```

# LAKE HURON

```{r}
LakeHuron
plot(LakeHuron,
     col='steel blue',
     frame=FALSE,
     main="Level of Lake Huron"
     )
```
Annual measurements of the level, in feet, of Lake Huron 1875–1972.


# TIME SERIES SIMULATION

We can sumulate times series using r command.

## White noise

```{r White-Noise}
n<-500
wn<-rnorm(n)
plot.ts(wn,
     col='steel blue',
     main='White noise',
     frame.plot = FALSE
     )
```

## Random walk

```{r Random-Walk}
n<-500
rwalk<-cumsum(rnorm(n))
plot.ts(rwalk,
     col='steel blue',
     main='Random walk',
     frame.plot = FALSE
     )
```

## Another random walk

```{r}
rwalk2<-cumsum(rnorm(n)+rnorm(n))
plot.ts(rwalk2,
     col='steel blue',
     main='random walk 2',
     frame.plot = FALSE
     )


```

## Cumulated random walk 

```{r}
rwalk3<-cumsum(cumsum(rnorm(n)))
plot.ts(rwalk3,
     col='steel blue',
     main='cumulated random walk',
     frame.plot = FALSE
     )

rwalk4<-cumsum(cumsum(rnorm(n)+rnorm(n)+rnorm(n)))
plot.ts(rwalk4,
     col='steel blue',
     main='cumulated random walk',
     frame.plot = FALSE
     )
```

## With a trend

```{r}
tseries<-1-(1:n)/100+cumsum(rnorm(n))+.2*rnorm(n)
plot.ts(tseries,
        col='steel blue',
        frame=FALSE
        )

```
```{r}
x<-(1:n)/100
tseries1<- .3 + .5*cos(2*pi*x) - 1.2*sin(2*pi*x) +
      .6*cos(2*2*pi*x) + .2*sin(2*2*pi*x) +
      -.5*cos(3*2*pi*x) + .8*sin(3*2*pi*x)+0.2*rnorm(n)
  
plot.ts(tseries1,
        col='steel blue',
        frame=FALSE,
        )
```

# AUTOCORRELATION

Ttme series observations are not independent, to check the dependence we can use the autocorrelation function. In R the autocorrelation is computed using the _acf_ function.

```{r, fig.height=12}
par(mfrow=c(4,2))
acf(LakeHuron,main='Lake Huron')
acf(wn,main='White noise')
acf(rwalk,main=names(rwalk))
acf(rwalk2,main=names(rwalk2))
acf(rwalk3,main=names(rwalk3))
acf(rwalk4,main=names(rwalk4))
acf(tseries,main=names(tseries))
acf(tseries1,main=names(tseries1))
```


```{r}
co2
plot.ts(co2,
        col="steel blue",
        frame=FALSE
        )
y<-diff(co2,lag=12)
x<-diff(co2)
par(mfrow=c(2,2))
plot.ts(y,col='steel blue')
plot.ts(x,col='steel blue')
acf(x)
acf(y)
```

