---
title: "TIME SERIES TP 1"
output:
  pdf_document:
    toc: yes
  html_document:
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      comment = NA,
                      warning = NA,
                      fig.width=12,
                      fig.height = 6
                      )
```
<style>
div.blue pre {background-color:#ccffcc;}
div.blue pre.r{background-color:#ccffff;}

</style>
<div class='blue'>

```{r}
library(ggplot2)
library(astsa)
library(dplyr)
library(xts)
```

# EXERCISE 1

## Question 1

Ths *ts*($data$, $start$=1, $end$=numeric, $frequency$=1,...) function is used to create  a time series object.

* $data$:  a vector or matrix of the observed time series

* $start$: the time of the first observation

* $end$: the time of the last observations , specified in the same way as start.

* $frequency$: the numeber of observations par unit of time

The *diff*(data)($x$, $lag$=1) returns suitably and iterated differences

* $x$: a numeric vector or a  matrix containing the values to be differenced

* $lag$: an integer indicating which lag to use

## Question 2

The *choose.file function choose a file interactively from the current repository.

Not very efficient when we use rmarkdown for a final report.

```{r}
if(!exists('euro50')) 
{
    euro50<-read.csv('tableEurom.csv',header = TRUE)
}
head(euro50)
```


## Question 3

The data contains different informations about a financial index  as: the prices at the opening and closing market, the lower and the higher price and so forth.

*euro50[,7]* contains the price at the closing time.

## Question 4

```{r,fig.height=12}
par(mfrow=c(2,2))
tseries50<-ts(euro50)
tslogdiff<-diff(log(tseries50[,7]))
plot(tseries50[,7],
     main='Adj.Close',
     col='steel blue',
     ylab='Adj.Close',
     frame=FALSE
    )
plot(tslogdiff,
     col='steel blue',
     main='difference of log Adj.Close',
     frame=FALSE
     )
hist(tslogdiff,
     col='steel blue',
     border = 'cyan',
     )
qqnorm(tslogdiff,
       col='steel blue',
       frame=FALSE,
       )
qqline(tslogdiff,
       col='red',
       lty=2,
       lwd=2
       )
```

# EXERCISE 2

## Question 1

```{r}
raindata<-read.table('pluie1.txt',
                     sep='',
                     header = TRUE
                     )
head(raindata)
tseriesrain<-ts(raindata$pluie,start = 1950,frequency = 12)
par(mfrow=c(1,2))
plot(tseriesrain,col='steel blue',main='Rain time series')
aggregate(tseriesrain,1,sum)%>%plot(col='steel blue',main='Cumulated')
```


The last figure shows a downward trend. We also notice the independence of aggregated data.

# EXERCISE


## Question 1


```{r,fig.height=12}
n<-1000
wn<-rnorm(n)
par(mfrow=c(2,2))
plot.ts(wn,col='steel blue',main='White noise')
acf(wn)
qqnorm(ts(wn),
       col='steel blue',
       frame=FALSE,
       )
qqline(ts(wn),
       col='red',
       lty=2,
       lwd=2
       )
hist(ts(wn),col='steel blue',border = 'cyan')
```


## Question 2

```{r}
shapiro.test(ts(wn))
```
The null hypothesis (observed data distribution is normal) can be reject. 

## Question 3

```{r,fig.height=12}
par(mfrow=c(2,3))
for(k in 1:6){rnorm(500)%>%ts()%>%acf()}
```

